Study on the effectiveness of the investment strategy based on a classifier with rules adapted by machine learning

### 2013-12-13 Article/Article20130912/LatexArt/tewi-paper-template_minipages.tex
aded: about PSO

The tests were deliberately performed in a fragment of the time series of a heavily diversified course, which contains both rising and downward trends as well as horizontal elements (Fig. \ref{timeSeries}). This time series, consists of 4734  \mbox{1-hour } candles, of the most important and the most fluent currency pair EURUSD from 22.10.2012.
\begin{figure}[h!]
\begin{center}
\includegraphics[width = 0.6\textwidth]{pictures/eurusd221012.png}
\end{center}
\caption{Time series EURUSD 1h}
\label{timeSeries}
\end{figure}
\FloatBarrier
A choice of parameters essential for defining the rules of opening and closing positions is crucial to the effectiveness of the strategy. Parametric space presented in this paper is a result of many trials prior to its final approval.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Characteristics of investment strategies}
The objective of the two strategies considered is to make investment decisions about buying or selling - opening long or short position in the market studied - the currency pair EURUSD.  The decision is based on the intersection of the current price and one of two barriers of additional indicator, called the ribbon. The band is made of two values calculated at the opening of each candle on the basis of historical data of the market. During the candle, values of the band do not change; therefore, barriers are creating step functions. In case when the current price exceeds any of the barrier values (goes out of the band), a decision to buy or sell is made - the type of decision depends on the variant of the considered strategy - decision for sub strategy TewiMiC is different than in case of TewiMiD. Names of the strategies are derived from the name of the project, in which the research was carried out.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Strategy analysis}
In both strategies a fixed period of learning are assumed (in the presented solution - 1000 one hour candles), followed by a testing period. Data from learning period were used to find a class of patterns which allowed to achieve maximum for the selected criterion, \textcolor{RedViolet}{in this study Calmar ratio (which is defined as a returned profit to maximum drawdown ratio) was selected.} The same patterns were then searched during the test period and the test results were computed for previously unused data space. Of course, these results do not have to already be positive and acceptable and could negatively surprise investors. During the test as a measure of the effectiveness of the investment was considered the maximum rate of net profit (with transaction costs). The authors believe that these two criteria in evaluation the quality of simulation results are legitimate. In the first phase of the validation, the training period is indicated for moderate and prudent risk management. In the test phase (in terms of actual trading) investor is mainly interested in profit.\\

@@

Given that the search space is relatively broad and unknown (it is difficult to estimate how specific combination of parameters would influence strategy's effectiveness) it is necessary to define its boundaries and then to find a combination of parameters that would maximize strategy's efficiency.  
In the first step of the process, a pseudo-random strategy is used to find the boundaries of the parameter space in which may exist the optimal solution. Calmar ratio value was used to assess the adequacy of randomly selected parameter combinations for a given period of learning. It would be possible to use an iterative method that would traverse the search space with certain step. However it would be extremely time consuming method given that the space is 11-dimensional. 
In the second stage PSO (Particle Swarm Optimization) \cite{Kennedy}  algorithm which already has been proved useful in finding optimal strategy parameters \cite{PSOtrading} was used, searching for solutions within prescribed limits of the first stage. The objective function was to maximizing the Calmar ratio, as it was in the first step. \\
The above tactic was used for each stage of the learning period, then checked "sustainability" of designated sets of parameters for test periods of different lengths - the basic rate of 100 candles and the other in the range of 10-400.
Figure ~\ref{research_methods} on page ~\pageref{research_methods} shows how the research was conducted. Having historical data for 1000 candles, optimal parameters, for said data, have been found using approach described above. After the parameters search, tests were performed on the current data. The strategy for the learned parameters should be used as long as it will bring satisfactory results on new data. When results were no longer good enough, the next parameters search were performed on next piece of historical data. Thus, the authors aim is to determine the point where those parameters should be recalculated. Additionally, the authors set out to test a new standard of quality prediction. Now, extending the period of testing can produce better results, but more slowly or with local drawndown in comparision to first period, when the classifier "remembers" the nature of the market. This new criteria is profit attributable to one candle of the testing period.\\

\begin{figure}[h!]
\begin{center}
\includegraphics[width=0.4\textwidth]{pictures/metodyka_badan.png}
\caption{Methods of testing carried out to test the variable size window}
\label{research_methods}
\end{center}
\end{figure}
\FloatBarrier

@@

\begin{figure}[h!]
\begin{minipage}{.49\linewidth}
\begin{center}
\includegraphics[width=0.9\textwidth]{pictures/mic_100.png}
\subcaption{Sub strategy TewiMiC}
\label{MiC100}
\end{center}
\end{minipage}
\begin{minipage}{.49\linewidth}
\begin{center}
\includegraphics[width=0.9\textwidth]{pictures/mid_100.png}
\subcaption{Sub strategy TewiMiD}
\label{MiD100}
\end{center}
\end{minipage}

@@

\begin{figure}[h!]
\begin{minipage}{.49\linewidth}
\begin{center}
\includegraphics[width=0.82\textwidth]{pictures/cumulativeReturnsC.png}
\subcaption{Results in 3D}
\label{Cum3DMiC}
\end{center}
\end{minipage}
\begin{minipage}{.49\linewidth}
\begin{center}
\includegraphics[width=0.82\textwidth]{pictures/mic_end.png}
\subcaption{Final results}
\label{Cum3DMiCend}
\end{center}
\end{minipage}
\caption{Earnings accumulated over time, depending on the length of the test period for TewiMIC}
\end{figure}
\FloatBarrier

@@

\begin{figure}[h!]
\centering
\begin{minipage}{.49\linewidth}
\begin{center}
\includegraphics[width=0.82\textwidth]{pictures/cumulativeReturnsPerCandleC.png}
\subcaption{Results in 3D}
\label{Cum3DPerCMiC}
\end{center}
\end{minipage}
\begin{minipage}{.49\linewidth}
\begin{center}
\includegraphics[width=0.82\textwidth]{pictures/mic_percandle_end.png}
\subcaption{Final results}
\label{Cum3DPerCMiCend}
\end{center}
\end{minipage}
\caption{Cumulative profit for one candle in time depending on the length of the testing period for TewiMiC}
\end{figure}
\FloatBarrier

@@

\textcolor{RedViolet}{Authors are aware of the fact that the parametric space is broad and choosing right ones is not a trivial task. It is not obvious that choosen and presented in this paper parameters are the best. There are no obvious sources that would suggest which parameters should be considered in investment strategies. To determine the parameters used in presented strategies an inductive method (??computional intelligence??) has been used --- after adding parameter the results were assessed and when they were acceptable the following parameter was added to the strategy. Even though selected parameters resulted in high efficiency, it does not mean that one should refrain from searching for better choice of parameters.}\\

>>>>

The tests were deliberately performed in a fragment of the time series of a heavily diversified course, which contains both rising and downward trends as well as horizontal elements (Fig. \ref{timeSeries}). This time series, consists of 4734  \mbox{1-hour } candles, of the most important and the most fluent currency pair EURUSD from 22.10.2012.
\begin{figure}[h!]
\begin{center}
\includegraphics[width = 0.55\textwidth]{pictures/eurusd221012.png}
\end{center}
\caption{Time series EURUSD 1h}
\label{timeSeries}
\end{figure}
\FloatBarrier
A choice of parameters essential for defining the rules of opening and closing positions is crucial to the effectiveness of the strategy. Parametric space presented in this paper is a result of many trials prior to its final approval.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Characteristics of investment strategies}
The objective of the two strategies considered is to make investment decisions about buying or selling - opening long or short position in the market studied - the currency pair EURUSD.  The decision is based on the intersection of the current price and one of two barriers of additional indicator, called the ribbon. The band is made of two values calculated at the opening of each candle on the basis of historical data of the market. During the candle, values of the band do not change; therefore, barriers are creating step functions. In case when the current price exceeds any of the barrier values (goes out of the band), a decision to buy or sell is made - the type of decision depends on the variant of the considered strategy - decision for sub strategy TewiMiC is different than in case of TewiMiD. Names of the strategies are derived from the name of the project, in which the research was carried out.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Strategy analysis}
In both strategies a fixed period of learning are assumed (in the presented solution - 1000 one hour candles), followed by a testing period. Data from learning period were used to find a class of patterns which allowed to achieve maximum for the selected criterion, \textcolor{RedViolet}{in this study Calmar ratio (which is defined as a final profit to maximum drawdown ratio) was selected.} The same patterns were then searched during the test period and the test results were computed for previously unused data space. Of course, these results do not have to already be positive and acceptable and could negatively surprise investors. During the test as a measure of the effectiveness of the investment was considered the maximum rate of net profit (with transaction costs). The authors believe that these two criteria in evaluation the quality of simulation results are legitimate. In the first phase of the validation, the training period is indicated for moderate and prudent risk management. In the test phase (in terms of actual trading) investor is mainly interested in profit.\\

@@

Given that the search space is relatively broad and unknown (it is difficult to estimate how specific combination of parameters would influence strategy's effectiveness) it is necessary to define its boundaries and then to find a combination of parameters that would maximize strategy's efficiency.  
In the first step of the process, a pseudo-random strategy is used to find the boundaries of the parameter space in which may exist the optimal solution. Calmar ratio value was used to assess the adequacy of randomly selected parameter combinations for a given period of learning. It would be possible to use an iterative method that would traverse the search space with certain step. However it would be extremely time consuming method given that the space is 11-dimensional. 
In the second stage PSO (Particle Swarm Optimization) \cite{Kennedy}  algorithm, which already has been proved useful in finding optimal strategy parameters \cite{PSOtrading} and allows to find a satisfying combination of parameters values in relatively short time even in broad search spaces, was used. Search space for PSO algorithm has been define in previous step. The objective function was to maximizing the Calmar ratio, as it was in the first step. \\
The above tactic was used for each stage of the learning period, then checked "sustainability" of designated sets of parameters for test periods of different lengths - the basic rate of 100 candles and the other in the range of 10-400.
Figure ~\ref{research_methods} on page ~\pageref{research_methods} shows how the research was conducted. Having historical data for 1000 candles, optimal parameters, for said data, have been found using approach described above. After the parameters search, tests were performed on the current data. The strategy for the learned parameters should be used as long as it will bring satisfactory results on new data. When results were no longer good enough, the next parameters search were performed on next piece of historical data. Thus, the authors aim is to determine the point where those parameters should be recalculated. Additionally, the authors set out to test a new standard of quality prediction. Now, extending the period of testing can produce better results, but more slowly or with local drawndown in comparision to first period, when the classifier "remembers" the nature of the market. This new criteria is profit attributable to one candle of the testing period.\\

\begin{figure}[h!]
\begin{center}
\includegraphics[width=0.6\textwidth]{pictures/metodyka_badan.png}
\caption{Methods of testing carried out to test the variable size window}
\label{research_methods}
\end{center}
\end{figure}
\FloatBarrier

@@

\begin{figure}[h!]
\begin{minipage}{.45\linewidth}
\begin{center}
\includegraphics[width=0.9\textwidth]{pictures/mic_100.png}
\subcaption{Sub strategy TewiMiC}
\label{MiC100}
\end{center}
\end{minipage}
\begin{minipage}{.45\linewidth}
\begin{center}
\includegraphics[width=0.9\textwidth]{pictures/mid_100.png}
\subcaption{Sub strategy TewiMiD}
\label{MiD100}
\end{center}
\end{minipage}

@@

\begin{figure}[h!]
\begin{minipage}{.45\linewidth}
\begin{center}
\includegraphics[width=0.80\textwidth]{pictures/cumulativeReturnsC.png}
\subcaption{Results in 3D}
\label{Cum3DMiC}
\end{center}
\end{minipage}
\begin{minipage}{.45\linewidth}
\begin{center}
\includegraphics[width=0.80\textwidth]{pictures/mic_end.png}
\subcaption{Final results}
\label{Cum3DMiCend}
\end{center}
\end{minipage}
\caption{Earnings accumulated over time, depending on the length of the test period for TewiMIC}
\end{figure}
\FloatBarrier

@@

\begin{figure}[h!]
\centering
\begin{minipage}{.45\linewidth}
\begin{center}
\includegraphics[width=0.80\textwidth]{pictures/cumulativeReturnsPerCandleC.png}
\subcaption{Results in 3D}
\label{Cum3DPerCMiC}
\end{center}
\end{minipage}
\begin{minipage}{.45\linewidth}
\begin{center}
\includegraphics[width=0.80\textwidth]{pictures/mic_percandle_end.png}
\subcaption{Final results}
\label{Cum3DPerCMiCend}
\end{center}
\end{minipage}
\caption{Cumulative profit for one candle in time depending on the length of the testing period for TewiMiC}
\end{figure}
\FloatBarrier

@@

\textcolor{RedViolet}{Authors are aware of the fact that the parametric space is broad and choosing right ones is not a trivial task. It is not obvious that choosen and presented in this paper parameters are the best. There are no obvious sources that would suggest which parameters should be considered in investment strategies. To determine the parameters utilized in presented strategies an iterative computation has been used --- after adding parameter the results were assessed and when they were acceptable the following parameter was added to the strategy. Even though selected parameters resulted in high efficiency, it does not mean that one should refrain from searching for better choice of parameters.}\\
